---
title: "bodycon Quick Start Guide"
author: "Julia Riley & Fonti Kar"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{bodycon}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

```{r hex-sticker, echo=FALSE, out.width='200px'}
library(here)
knitr::include_graphics(
  here::here("inst", "hex_sticker", "hex_sticker.png"))
```

*bodycon* is an R package for calculating commonly-used body condition indices in wildlife ecology research. In this research field, an individual's fitness traits, like reproduction and survival, are the primary metrics used to quantify an animal's health, biology, and evolution. Yet, direct fitness measures like these can be un unfeasible to collect in the field due to financial or logistical research constraints. So, in these cases, animal ecologists often rely on proxies of individual fitness, such as body condition. In general, individuals in better body condition (that are also assumed to have greater energy reserves) experience higher survival and reproductive success than individuals in inferior body condition. This impact of body condition on fitness parameters has been documented across a diversity of taxa. One widely used method of estimating body condition is to relate body mass to a linear measure of body size, but this can be calculated in a variety of ways. This R package contains functions to for commonly-used methods of calculation of body condition indices; and we hope to continue to develop this R package to expand the taxa it can be applied to moving forward.

# Installation

To install from CRAN:
```{r}
install.packages("bodycon")
```

Or install the development version from GitHub:
```{r}
install.packages("remotes")
remotes::install_github("julia-riley/bodycon")
```

Load the package:
```{r}
library(bodycon)
```


# Package Datasets

There are two datasets included in the bodycon R package. The first contains data on the morphology of 46 Maritime Gartersnakes (**Thamnophis sirtalis pallidulus**) collected in summer of 2022 in New Brunswick and Nova Scotia, Canada. The length measure is snout-vent-length (mm), which is the distance from the snake's snout to the anterior edge of their cloaca. The weight measure is mass (g).
```{r}
# View the first (Maritime Gartersnake) dataset
glimpse(gartersnake)
```

There is also a second dataset that contains data on the morphology of 128 Eastern Red-backed Salamanders (**Plethodon cinereus**) that was collected the the summer of 2022 in New Brunswick, Canada. It also contains similar length (snout-vent-length in mm) and mass data for each individual salamander that can be used to calculate body condition indexes.
```{r}
# View the second (Eastern Red-backed Salamander) dataset
glimpse(salamander)
```


# Calculation of body condition indexes

There is one function `bci`in this R package you can use to calculate the body condition of animals. You can specify the method used to calculate body condition within this function. There are three possible methods. All rely on a standardized length measurement (i.e., tarsus length for birds or snout-vent-length for reptiles) and a measure of weight. 

## 1) Body condition index estimation using residuals from an OLS regression

This example calculates body condition indices from the residuals of an ordinary least squares regression (OLS). This is a traditional approach in ecology as outlined by \insertCite{krebs1993;textual}{bodycon}. The method to specify for this is `"resid_ols"` and please see the examples below of using it to calculate body condition of gartersnakes:

```{r}

bci_resid_ols(gartersnake, svl_mm, mass_g, method = "resid_ols")

# Piped version
gartersnake |>
  bci_resid_ols(svl_mm, mass_g, method = "resid_ols")

```


## Scaled mass body condition index (SMI) estimation

Yet, calculating body condition indices from the residuals of an ordinary least squares regression (OLS) has been critiqued in the literature. There is discussion whether it is the most robust approach \insertCite{schulte2005,peig2009}{bodycon}, how its appropriateness may vary by taxon \insertCite{jakob1996,buancilua2010,labocha2012}{bodycon}, and whether or not it fits the assumptions of certain statistical tests \insertCite{garcia2001}{bodycon}. An alternative is the scaled mass index (SMI method) as described in \insertCite{peig2009;textual}{bodycon}, which can be estimated using either a OLS or robust regression (as described in [this code by Chen-Pan Liao](https://apansharing.blogspot.com/2018/05/an-r-function-olsrobust-caled-mass-index.html)).

### With OLS regression

Below is an example of using the SMI method's function using the ordinary least squares (OLS) regression in its estimation of the body condition indices for salamanders. Yet, this method is sensitive to the presence of outliers (i.e., data points that may distort the expected relationship between body length and weight), and so SMI estimation using robust regression (see function `bci_smi_rob`) may be more appropriate in cases where outliers are present.

```{r}

bci_smi_ols(salamander, svl_mm, mass_g, method = "smi_ols")

# Piped version
salamander |>
  bci_smi_ols(svl_mm, mass_g, method = "smi_ols")

```

### With robust regression

Below is an example of using the SMI method's function using the robust regression using an M estimator (from the MASS R package) \insertCite{venables2002}{bodycon} in its estimation of the body condition indices for salamanders. This method is less sensitive to the presence of outliers (i.e., data points that may distort the expected relationship between body length and weight), as shown in [this blog by by Chen-Pan Liao](https://apansharing.blogspot.com/2018/05/an-r-function-olsrobust-caled-mass-index.html).

```{r}

bci_smi_rob(salamander, svl_mm, mass_g, method = "smi_rob")

# Piped version
salamander |>
  bci_smi_rob(svl_mm, mass_g, method = "smi_rob")

```


## Using multiple methods at once

It is possible to estimate body condition indexes using multiple methods. This may be something you might like to do to decide on which method you are going to use for your data. Here is an example of how you can do that:

```{r}

bci_resid_ols(gartersnake, svl_mm, mass_g, 
              method = c("resid_ols", "smi_ols", "smi_rob"))

# Piped version
gartersnake |>
  bci_resid_ols(svl_mm, mass_g, 
                method = c("resid_ols", "smi_ols", "smi_rob"))

```


# Visual comparison of body condition indices

To explore the body condition methods that are most appropriate for you data, you may want to plot the estimations with your observed, raw data to visually compare methods. Of course, please also consider currently literature about these BCI estimators to assess whether it is correct for your dataset or the analyses you are planning to apply to them. We will plot BCI estimators for animals using the `plot_bci()` function from the bodycon package. The plot allows comparison between:

+ OLS residuals (resid_ols)
+ Scaled Mass Index using OLS regression (smi_ols)
+ Scaled Mass Index using robust regression (smi_rob)

Additionally, raw data points can be grouped (e.g., by sex or population) and colours can be customized for both raw points and method lines. In the following examples we will use the `gartersnakes` dataset, which has previously been outlined above.

### Basic Plot
An example of a simple plot of all three BCI methods with default colours.

```{r}
plot_bci(
   gartersnake,
   svl_mm,
   mass_g,
   methods = c("resid_ols", "smi_ols", "smi_rob")
   )
```

### Grouped plot
You can highlight different groups in the raw data using the `group` argument. For example, here we separate male and female gartersnakes and the fit of the SMI method using robust regression:

```{r}  
plot_bci(
   gartersnake,
   svl_mm,
   mass_g,
   methods = c("smi_rob"),
   group = sex
   )
```

# Custom colours and a hidden legend
You can further adjust your plot with custom colours for the observed data groups, the methods lines, and adjust whether or not there is a legend: 
```{r}  
plot_bci(
   gartersnake,
   svl_mm,
   mass_g,
   methods = c("smi_rob"),
   group = sex,
   raw_colours = c("M" = "darkorange3", "F" = "darkorchid"),
   method_colours = c("SMI (robust)" = "black")
)
```

# References

Băncilă RI, Hartel T, Plăiaşu R, Smets J, and Cogălniceanu D. (2010) Comparing three body condition indices in amphibians: a case study of yellow-bellied toad *Bombina variegata*. Amphibia-Reptilia, 31(4), 558-562.

García-Berthou E. (2001) On the misuse of residuals in ecology: testing regression residuals vs. the analysis of covariance. Journal of Animal Ecology, 70(4), 708-711.

Jakob EM, Marshall SD, and Uetz GW. (1996) Estimating fitness: a comparison of body condition indices. Oikos, 77, 61-67.

Krebs CJ, and Singleton GR. (1993) Indexes of condition for small mammals. Australian Journal of Zoology, 41(4), 317-323.

Labocha MK, and Hayes JP. (2012) Morphometric indices of body condition in birds: a review. Journal of Ornithology, 153(1), 1-22.

Peig J, and Green AJ. (2009) New perspectives for estimating body condition from mass/length data: the scaled mass index as an alternative method. Oikos, 118(12), 1883-1891.

Schulte-Hostedde AI, Zinner B, Millar JS, and Hickling GJ. (2005) Restitution of mass–size residuals: validating body condition indices. Ecology, 86(1), 155-163.
